Social Participation Ontology: community documentation, enhancements and use examples

### 2015-01-10 paper.tex
retoque de escrita

\author{Rodrigo Bandeira de Luna}
  \homepage{http://www.cidadedemocratica.org.br/}
  \email{rodrigoyellow@cidadedemocratica.org.br}
 \altaffiliation{Cidade Democr\'atica}%Lines break automatically or can be forced with \\
%
%\author{Vilson Vieira da Silva Junior}
%  \homepage{http://automata.cc/}
%  \email{vilson@void.cc}
%  \altaffiliation[Also at ]{IFSC-USP}%Lines break automatically or can be forced with \\
%
%\author{Ricardo Fabbri}
%  \homepage{http://www.lems.brown.edu/~rfabbri/}
%  \email{rfabbri@iprj.uerj.br}
% \altaffiliation{
%Instituto Polit\'ecnico, Universidade Estadual do Rio de Janeiro (IPRJ)
%}%Lines break automatically or can be forced with \\
%
%\author{Deborah Christina Antunes}
%  \homepage{http://lattes.cnpq.br/1065956470701739}
%  \email{deborahantunes@gmail.com}
%  \altaffiliation{
%Curso de Psicologia, Universidade Federal do Cer\'a (UFC - Sobral)
%}%Lines break automatically or can be forced with \\
%
%\author{Marilia Mello Pisani}
%  \homepage{http://lattes.cnpq.br/6738980149860322}
%  \email{marilia.m.pisani@gmail.com}
% \altaffiliation{
%Centro de Ci\^encias Naturais e Humanas, Universidade Federal do ABC (CCNH/UFABC)
%}
%
%\author{Luciano da Fontoura Costa}
%  \homepage{http://cyvision.ifsc.usp.br/~luciano/}
%  \email{ldfcosta@gmail.com}
%  \altaffiliation[Also at ]{IFSC-USP}%Lines break automatically or can be forced with \\
%
%\author{Ricardo Augusto Poppi Martins}
%  \homepage{http://ricardopoppi.wordpress.com/}
%  \email{ricardo.poppi@presidencia.gov.br}
% \altaffiliation{Secretaria-Geral da Presidência da República (SG/PR)}%Lines break automatically or can be forced with \\
%
%\author{Ronald Emerson Scherolt da Costa}
%  \homepage{http://www.icmc.usp.br/pessoas/dilvan/}
%  \email{ronald.costa@presidencia.gov.br}
% \altaffiliation[Also at ]{SG/PR}%Lines break automatically or can be forced with \\
%
%
%\author{Osvaldo Novais de Oliveira Junior}
%  \homepage{www.polimeros.ifsc.usp.br/professors/professor.php?id=4}
%  \email{chu@ifsc.usp.br}
% \altaffiliation[Also at ]{IFSC-USP}%Lines break automatically or can be forced with \\
%

@@

%\newcommand{\t}[1]{\texttt{#1}}
\begin{abstract}
    Participatory democracy advances in virtually all governments. South America presents a prominent context with mixed culture and social predisposition. At 2012, civil, academic and governmental parties started elaborating the ``Common Vocabulary of Social Participation'' (\vcps\ from the Brazilian name \emph{Vocabul\'ario Comum de Participa\c{c}\~ao Social}), as a public and online process. By May, 2013, first reference documents were publicized, together with a preliminary \owl\ code,
logos, and a diagram of a general ``public consultation''.
The Corais Platform kept online records of the process, like discussions and preparation of texts. 
This article exposes this material and proposes elementary unfolding, highlighting the ``Social Participation Ontology'' (\ops\ from the Brazilian name \emph{Ontologia de Participa\c{c}\~ao Social}). To exhibit this new ontology, this steps were considered: completion of \vcps\ \owl\ code for preliminary \ops, with necessary translations and corrections; correction of ontological contradictions and enhancements of class names and labels in Portuguese, Spanish and English; a toy expansion of the ontology by further specifying classes; use cases of \ops\ by researchers and public managers; and use examples regarding knowledge organization and a \sparql\ endpoint. Ongoing work involves incorporation by the official Brazilian social participation federal portal, incorporation by civil participative organizations, linkage to other participative ontologies. \ops\ is being used as an upper ontology, and all classes linked to \foaf\ and \bfo\ as higher upper ontologies.
\end{abstract}

@@

\begin{acknowledgments}
This article is deeply influenced by Protege and BFO documentation. Authors thank community and researchers related to these projects~\cite{protege,bfo}.
Renato Fabbri is grateful to CNPq (project 870336/1997-5), PNUD/ONU (contract 2013/00056, project BRA/12/018), SNAS/SG-PR, and the Postgraduate Committee of the IFSC/USP. Authors thank \corais\ Platform maintainers for their efforts in delivering a collaborative platform which gave birth to \vcps\ documentation. Authors thanks Prof. Dr. Dilvan de Abreu Moreira (ICMC/USP) for all help and incentives in the genesis of this article. Authors thank Flor Karina Mamani Amanqui for initial Spanish labels.
% 2013/00056, projeto BRA/12/018 (pelo que?? nao me lembro mais e nao anotamos)
\end{acknowledgments}

>>>>

\author{Rodrigo Bandeira de Luna}
  \homepage{http://www.cidadedemocratica.org.br/}
  \email{rodrigoyellow@cidadedemocratica.org.br}
 \altaffiliation{Cidade Democr\'atica}%Lines break automatically or can be forced with \\

\author{Ricardo Augusto Poppi Martins}
  \homepage{http://ricardopoppi.org/}
  \email{ricardo.poppi@presidencia.gov.br}
  \altaffiliation{Social Participation Department, Social Articulation National Secretariat, Brazilian Presidency General Secretariat of the Republic (DPS/SNAS/SG-PR)}%Lines break automatically or can be forced with \\

%
%\author{Vilson Vieira da Silva Junior}
%  \homepage{http://automata.cc/}
%  \email{vilson@void.cc}
%  \altaffiliation[Also at ]{IFSC-USP}%Lines break automatically or can be forced with \\
%
%\author{Ricardo Fabbri}
%  \homepage{http://www.lems.brown.edu/~rfabbri/}
%  \email{rfabbri@iprj.uerj.br}
% \altaffiliation{
%Instituto Polit\'ecnico, Universidade Estadual do Rio de Janeiro (IPRJ)
%}%Lines break automatically or can be forced with \\
%
%\author{Deborah Christina Antunes}
%  \homepage{http://lattes.cnpq.br/1065956470701739}
%  \email{deborahantunes@gmail.com}
%  \altaffiliation{
%Curso de Psicologia, Universidade Federal do Cer\'a (UFC - Sobral)
%}%Lines break automatically or can be forced with \\
%
%\author{Marilia Mello Pisani}
%  \homepage{http://lattes.cnpq.br/6738980149860322}
%  \email{marilia.m.pisani@gmail.com}
% \altaffiliation{
%Centro de Ci\^encias Naturais e Humanas, Universidade Federal do ABC (CCNH/UFABC)
%}
%
%\author{Luciano da Fontoura Costa}
%  \homepage{http://cyvision.ifsc.usp.br/~luciano/}
%  \email{ldfcosta@gmail.com}
%  \altaffiliation[Also at ]{IFSC-USP}%Lines break automatically or can be forced with \\
%
%\author{Ricardo Augusto Poppi Martins}
%  \homepage{http://ricardopoppi.wordpress.com/}
%  \email{ricardo.poppi@presidencia.gov.br}
% \altaffiliation{Secretaria-Geral da Presidência da República (SG/PR)}%Lines break automatically or can be forced with \\
%
%\author{Ronald Emerson Scherolt da Costa}
%  \homepage{http://www.icmc.usp.br/pessoas/dilvan/}
%  \email{ronald.costa@presidencia.gov.br}
% \altaffiliation[Also at ]{SG/PR}%Lines break automatically or can be forced with \\
%
%
%\author{Osvaldo Novais de Oliveira Junior}
%  \homepage{www.polimeros.ifsc.usp.br/professors/professor.php?id=4}
%  \email{chu@ifsc.usp.br}
% \altaffiliation[Also at ]{IFSC-USP}%Lines break automatically or can be forced with \\
%

@@

%\newcommand{\t}[1]{\texttt{#1}}
\begin{abstract}
    Participatory democracy advances in virtually all governments. South America presents a prominent context with mixed culture and social predisposition. In 2012, civil, academic and governmental parties started elaborating the ``Common Vocabulary of Social Participation'' (\vcps\ from the Brazilian name \emph{Vocabul\'ario Comum de Participa\c{c}\~ao Social}), as a public and online process. By May, 2013, first reference documents were publicized, together with a preliminary \owl\ code,
logos, and a diagram of a general ``public consultation''.
The \corais\ platform kept online records of the process, like discussions and preparation of texts. 
This article exposes this material and proposes an elementary unfolding: the ``Social Participation Ontology'' (\ops\ from the Brazilian name \emph{Ontologia de Participa\c{c}\~ao Social}). To exhibit this new ontology, these steps were considered: correction of ontological contradictions and \owl\ protocol use errors; completion of \vcps\ \owl\ code; translations and standardizations; enhancements of class names and labels in Portuguese, Spanish and English; a toy expansion of the ontology by further specifying classes; linked data examples regarding dereferencing, a \sparql\ endpoint and participatory instances; use cases by researchers and public managers. Ongoing work involves further adoption of \ops\ by the official Brazilian social participation federal portal, further adoption by civil participative organizations, and linkage to other participative ontologies. \ops\ is being used as an upper ontology, and all classes linked further to \foaf\ and \bfo\ as higher upper ontologies.
\end{abstract}

@@

\begin{acknowledgments}
This article is deeply influenced by Protege and BFO documentation. Authors thank community and researchers related to these projects~\cite{protege,bfo}.
Renato Fabbri is grateful to CNPq (process 140860/2013-4, project 870336/1997-5), PNUD/ONU (contract 2013/00056, project BRA/12/018), SNAS/SG-PR, and the Postgraduate Committee of the IFSC/USP. Authors thank \corais\ Platform maintainers for their efforts in delivering a collaborative platform which gave birth to \vcps\ documentation. Authors thanks Prof. Dr. Dilvan de Abreu Moreira (ICMC/USP) for all help and incentives in the genesis of this article. Authors thank Flor Karina Mamani Amanqui for initial Spanish labels.
% 2013/00056, projeto BRA/12/018 (pelo que?? nao me lembro mais e nao anotamos)
\end{acknowledgments}
